What is an Ancestral Recombination Graph?

### 2022-02-15 paper.tex
Attempts at definitions.

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format.
The concrete details of how an ARG should be stored and processed
are not discussed in the literature, % weak, fix.
but the following is a reasonable approach that mirrors the implementations
in methods such as [x, y z].
A Griffiths ARG is composed of a set of $N$ nodes, where each node is an
integer $0 \leq u < N$, and edges $(u, v)$ define the relationships between
those nodes (Fig.~\ref{fig-arg-data-structure}A).
Each node corresponds to an \emph{event}, and is either a common ancestor
event (in which case it has two children and one parent)
or a recombination event (when it has one child and two parents).
Each recombination event is associated with a breakpoint $0 < x < L$,
which we might encode as a mapping $f: N \rightarrow \{0, ..., L - 1\}$
such that $f(u)$ is the recombination breakpoint if $u$ is a recombination
node and $f(u) = \infty$ otherwise. This encoding captures the information
of a realisation of the ARG stochastic process, but it is important to
note that it does not uniquely describe the underlying ancestral history
without stipulating some additional ordering requirements.

@@

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
The most important change between the two encodings is to represent
a recombination event (node 3 in A) by two parent genomes
(nodes 3 and 4 in B) and explicitly storing the interval
that the child inherits from each parent along the two corresponding
edges. Thus, child $1$ in inherits $(0, x]$ from $3$
and $(x, L)$ from $4$
(for recombination breakpoint $x$ and a sequence of length $L$).
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

\section*{Anchoring the ARG to the pedigree}

The deep connection between the ARG and the pedigree is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions of the ARG have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

@@

\section*{Ancestry resolution and simplification}
% Note: we want some terminology here to be precise about what we're
% talking about - perhaps a G-ARG for genome-ARG vs E-ARG for event?
% Just using ARG for now, can clean up in another pass.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

>>>>

The Griffiths ARG encoding is an elegant and mathematically
economical description of the coalescent with recombination on a single
chromosome, but it has significant drawbacks as the basis for a general purpose
storage and interchange format. To avoid confusion, we will refer to this
ARG encoding as an ``Event ARG'', or EARG.
There are many ways in which the EARG structure can be formalised (see
e.g.~\citep{gusfield2014recombinatorics}).
% It would be slightly easier for us if we defined nodes as in N_0, since
% we use 0 as a node ID, but leaving for now.
We can define an EARG as a tuple $(e, \sigma)$, where $e$ is an ordered
list of tuples such that $e_j = (c, p)$ defines an edge between
child node $c$ and parent $p$ ($c, p \in \mathbb{N}$), and
$\sigma: \mathbb{N} \rightarrow \mathbb{N}_0$ is a function
mapping nodes to recombination breakpoints, such that $f(u) = x$
if $u$ is a recombination event with breakpoint $x$ and
% Because of our definitions we can't have anything to the left of 0,
% so you can treat breakpoint at 0 as no recombination.
$f(u) = 0$ otherwise. Note that this description captures only the
graph topology: if we also wish to know the times of events we need
an additional function $t(u): \mathbb{N} \rightarrow \mathbb{R}$
which defines the time of each node.
An example is given in Fig.~\ref{fig-arg-data-structure}A.

@@

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
[TODO clarify that the function is the same things as having two
types of nodes]
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
The most important change between the two encodings is to represent
a recombination event (node 3 in A) by two parent genomes
(nodes 3 and 4 in B) and explicitly storing the interval
that the child inherits from each parent along the two corresponding
edges. Thus, child $1$ in inherits $(0, x]$ from $3$
and $(x, L)$ from $4$
(for recombination breakpoint $x$ and a sequence of length $L$).
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

\section*{Genome ARGs}

The deep connection between ARGs and pedigrees is of course
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of the ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the expressivity of the structure.
Following \cite{mathieson2020ancestry} we instead define the
ARG data structure \emph{in terms of} an underlying pedigree, so that
any form of inheritance structure can be modelled.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

@@

Formally, we can define a GARG as a set of edges $E$, where each
element of the set is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r])$ over which genome $c$ inherits from $p$ (we may
also phrase this in terms of the ancestral material of a sample---see
the next section).

@@

\section*{Ancestry resolution and simplification}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

Formally, we can define a sample-resolved GARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r])$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.
